---
title: "Tennis Match Outcome Predictions"
author: "Team: SNM Analytics"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

---
title: "Predicting Tennis Match Outcomes: A Data-Driven Approach to the Game"
author: "Team: SNM Analytics"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---



```{r init, include=F}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
options(scientific = T, digits = 3)
# Loading libraries
library(ezids)
library(dplyr)
library(tidyverse)
library(readr)
library(ggplot2)
```

# Introduction

## Project Overview

In this analysis, we will examine the Jeff Sackmann tennis ATP match dataset to determine the availability of match-level statistics over the years. The dataset contains information on tennis matches from 1968 to 2023, with more extensive match statistics available starting from 1997. Our goal is to visualize the availability of match statistics and explore the data to understand its structure and contents.

## Our SMART Question:

What are the most important features that influence the outcome of a tennis match, and how accurately can we predict the winner of a match using a machine learning model based on these features?

This question not only focuses on prediction accuracy but also emphasizes the importance of understanding which factors contribute the most to the outcome. By answering this question, you can gain insights into the game and potentially develop strategies for players or coaches to improve performance.


```{r}
# Initialize an empty data frame
data <- data.frame()

# Loop through the years and read the corresponding atp_matches files
for (year in 1968:2023) {
  file_path <- paste0("data/atp_matches_", year, ".csv")
  if (file.exists(file_path)) {
    temp_data <- read.csv(file_path)
    data <- rbind(data, temp_data)
  }
}
```


```{r}
str(data)
```


## Dataset Description

The dataset contains information about tennis matches played between 1963 and 2023, with a total of 189,185 records. It includes 49 columns, which provide information about the players, match statistics, and other relevant details.

Some of the key features in the dataset are:

1. Player details: player names, ages, heights, etc.
2. Match details: tournament name, surface, round, etc.
3. Player rankings: ATP rankings and ranking points.
4. Match statistics: aces, double faults, first serves in, first serve points won, etc.


```{r}
# Convert tourney_date character type to a Date object
data$tourney_date <- as.Date(as.character(data$tourney_date), "%Y%m%d")
```


```{r}
# Extract year from tourney_date
data$year <- year(data$tourney_date)
```


```{r}
# Select the relevant columns (numeric match statistics)
stats_columns <- c("w_ace", "w_df", "w_svpt", "w_1stIn", "w_1stWon", "w_2ndWon", "w_SvGms", "w_bpSaved", "w_bpFaced",
                   "l_ace", "l_df", "l_svpt", "l_1stIn", "l_1stWon", "l_2ndWon", "l_SvGms", "l_bpSaved", "l_bpFaced")

```

```{r}
# Calculate the percentage of non-null values per year for each numeric variable
non_null_stats <- data %>%
  group_by(year) %>%
  summarize(across(all_of(stats_columns), ~mean(!is.na(.)) * 100, .names = "pct_non_null_{col}"))
```

```{r}
# Gather the data into a long format for easy plotting
non_null_stats_long <- non_null_stats %>%
  gather(key = "variable", value = "pct_non_null", -year)
```

```{r}
# Plot the percentage of non-null values per year for each numeric variable
ggplot(non_null_stats_long, aes(x = year, y = pct_non_null, color = variable)) +
  geom_line() +
  geom_point() +
  labs(title = "Percentage of Non-Null Match Statistics per Year",
       x = "Year",
       y = "Percentage of Non-Null Values",
       color = "Statistic") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 8)) +
  scale_x_continuous(breaks = seq(min(non_null_stats_long$year), max(non_null_stats_long$year), by = 5))


```


```{r}
library(knitr)

kable(non_null_stats, digits = 2)

```

The Jeff Sackman Dataset contains multiple sets of files. We use the Sackman_tennis_atp-master project which contains match level
statistics from the year 1968. Since more extensive match level statistics are only available from the year 1991, we concentrate on the ATP match level files from 1991 to 2023.

```{r}
data <- data %>% filter(year >= 1991)
```


```{r}
str(data)
```

# Data Preprocessing & Cleaning

```{r}
# Analyze missing values in the dataset
missing_values <- data %>%
  gather(variable, value, -tourney_id, -tourney_name, -tourney_date, -match_num, -winner_id, -loser_id) %>%
  mutate(is_missing = is.na(value)) %>%
  group_by(variable) %>%
  summarize(missing_count = sum(is_missing), total_count = n(), missing_percentage = missing_count / total_count * 100)

print(missing_values)
```


```{r}
# Analyze missing values by year
missing_values_by_year <- data %>%
  gather(variable, value, -tourney_id, -tourney_name, -tourney_date, -match_num, -winner_id, -loser_id, -year) %>%
  mutate(is_missing = is.na(value)) %>%
  group_by(year, variable) %>%
  summarize(missing_count = sum(is_missing), total_count = n(), missing_percentage = missing_count / total_count * 100) 

print(missing_values_by_year)
```


Renamed the Variables for better understanding

```{r}
data <- data %>% 
  rename(tournament_id = tourney_id,
         tournament_name = tourney_name,
         tournament_level = tourney_level,
         tournament_date = tourney_date,
         winner_height = winner_ht,
         winner_country = winner_ioc,
         loser_height = loser_ht,
         loser_country = loser_ioc,
         winner_aces = w_ace,
         winner_doublefaults = w_df,
         winner_servepointswon = w_svpt,
         winner_first_serve_in = w_1stIn,
         winner_first_serve_won = w_1stWon,
         winner_second_serve_won = w_2ndWon,
         winner_service_games_held = w_SvGms,
         winner_breakpoints_saved = w_bpSaved,
         winner_breakpoints_faced = w_bpFaced,
         loser_aces = l_ace,
         loser_doublefaults = l_df,
         loser_servepointswon = l_svpt,
         loser_first_serve_in = l_1stIn,
         loser_first_serve_won = l_1stWon,
         loser_second_serve_won = l_2ndWon,
         loser_service_games_held = l_SvGms,
         loser_breakpoints_saved = l_bpSaved,
         loser_breakpoints_faced = l_bpFaced)
```


```{r}
summary(data)
```

```{r}
# Define a custom color palette for the surfaces
surface_colors <- c("Clay" = "#EB9694", 
                    "Carpet" = "#FABE58", 
                    "Grass" = "#87D37C", 
                    "Hard" = "#5DADE2")

# Create the bar chart with custom colors and adjusted font sizes
ggplot(data, aes(x = surface, fill = surface)) + 
  geom_bar(color = "black") +
  scale_fill_manual(values = surface_colors) +
  labs(title = "Total Matches Played on Various Surfaces Since 1991",
       x = "Surface",
       y = "Total Matches") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10), size = 12),
        axis.text = element_text(size = 12),
        panel.grid.major = element_line(colour = "gray90"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 18, hjust = 0.5, color = "gray40"),
        plot.caption = element_text(size = 14, hjust = 0, color = "gray40"),
        plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

```

Most of the matches played are on hard courts!

```{r}
# Filter out rows with missing winner_age and loser_age values
filtered_data <- data %>%
  filter(!is.na(winner_age) & !is.na(loser_age))

# Create a density plot for winner_age and loser_age
ggplot() +
  geom_density(data = filtered_data, aes(x = winner_age, fill = "Winner"), alpha = 0.5) +
  geom_density(data = filtered_data, aes(x = loser_age, fill = "Loser"), alpha = 0.5) +
  labs(x = "Age", y = "Density", title = "Age Distribution of Winners and Losers", fill = "Role") +
  theme_minimal() +
  theme(legend.position = "top")

# Create a box plot for winner_age and loser_age
winner_loser_age <- filtered_data %>%
  select(winner_age, loser_age) %>%
  gather(key = "Role", value = "Age", winner_age, loser_age)

ggplot(winner_loser_age, aes(x = Role, y = Age, fill = Role)) +
  geom_boxplot() +
  labs(x = "", y = "Age", title = "Age Distribution of Winners and Losers") +
  theme_minimal() +
  theme(legend.position = "none")
```
age distribution for "Losers" is wider than "Winners", although mean value is quite the same: younger and olders players are more likely to loose matches.


```{r}

# Filter out rows with missing winner_age and loser_age values
filtered_data <- data %>%
  filter(!is.na(winner_aces) & !is.na(loser_aces))

# Create a density plot for winner_ace and loser_ace
ggplot() +
  geom_density(data = filtered_data, aes(x = winner_aces, fill = "Winner")) +
  geom_density(data = filtered_data, aes(x = loser_aces, fill = "Loser")) +
  labs(x = "Ace", y = "Density", title = "Ace Distribution of Winners and Losers", fill = "Role") +
  theme_minimal() +
  theme(legend.position = "top")


```

Ace Distribution between losers and winners.

```{r}
str(data)
```

```{r}
filtered_data <- data %>%
  mutate(winner_first_serve_pct = winner_first_serve_won / winner_first_serve_in,
         winner_second_serve_pct = winner_second_serve_won / (winner_servepointswon - winner_first_serve_in),
         loser_first_serve_pct = loser_first_serve_won / loser_first_serve_in,
         loser_second_serve_pct = loser_second_serve_won / (loser_servepointswon - loser_first_serve_in))
```

```{r}
filtered_data
```

```{r}
# Prepare the data
winner_loser_data <- filtered_data %>%
  select(winner_first_serve_pct, loser_first_serve_pct) %>%
  gather(key = "Role", value = "First Serve Win Percentage", winner_first_serve_pct, loser_first_serve_pct)

# Create the box plot
box_plot <- ggplot(winner_loser_data, aes(x = Role, y = `First Serve Win Percentage`, fill = Role)) +
  geom_boxplot() +
  labs(x = "", y = "First Serve Win Percentage") +
  theme_minimal() +
  theme(legend.position = "none")

# Display the box plot
print(box_plot)

```


As expected winner first serve winning percentage is quite higher

```{r}
str(data)
```

```{r}
data = data[sample(1:nrow(data)), ]
```


```{r}
str(data)
```



```{r}
# Determine the number of rows in half the dataset
half_nrow <- nrow(data) %/% 2

# Split the dataset in half
first_half <- data[1:half_nrow, ]
second_half <- data[(half_nrow + 1):nrow(data), ]

# Rename columns for the first half
colnames(first_half)[grep("winner", colnames(first_half))] <- gsub("winner", "player0", colnames(first_half)[grep("winner", colnames(first_half))])
colnames(first_half)[grep("loser", colnames(first_half))] <- gsub("loser", "player1", colnames(first_half)[grep("loser", colnames(first_half))])

# Rename columns for the second half
colnames(second_half)[grep("winner", colnames(second_half))] <- gsub("winner", "player1", colnames(second_half)[grep("winner", colnames(second_half))])
colnames(second_half)[grep("loser", colnames(second_half))] <- gsub("loser", "player0", colnames(second_half)[grep("loser", colnames(second_half))])

# Add 'winner' column to the first half and set it to 0 (player0 is the winner)
first_half$winner <- 0

# Add 'winner' column to the second half and set it to 1 (player1 is the winner)
second_half$winner <- 1

# Combine the two halves
renamed_data <- rbind(first_half, second_half)
```


```{r}
str(renamed_data)
```

```{r}
# Replace 'tennis_data' with the name of your data frame
# write.csv(renamed_data, "tennis_data.csv", row.names = FALSE)
```

```{r}
final_data = renamed_data
```

```{r}
ncol(final_data)
```

```{r}
final_data = read_csv("tennis_data.csv", show_col_types = FALSE)
```

```{r}
library(corrplot)
```

```{r}
numeric_vars <- final_data[, sapply(final_data, is.numeric)]
correlation_matrix <- cor(numeric_vars, use = "complete.obs")
png("heatmap_full.png", width = 800, height = 800)
corrplot(correlation_matrix, method = "color", type = "upper", order = "hclust", tl.col = "black", tl.srt = 45)
dev.off()
```


```{r}
tennis_data_clean <- final_data %>%
  select(-tournament_id, -tournament_name, -tournament_date, -year, -player0_seed, -player1_seed, -player0_id, -player1_id, -player0_name, -player1_name, -match_num, -score, -round, -draw_size, -player1_country, -player0_country, -player0_entry, -player1_entry, -minutes)
```

```{r}
ncol(tennis_data_clean)
```

```{r}
unique(tennis_data_clean$player0_hand)
```


```{r}
surface_colors <- c("Clay" = "#EB9694", 
                    "Carpet" = "#FABE58", 
                    "Grass" = "#87D37C", 
                    "Hard" = "#5DADE2")
ggplot(data = tennis_data_clean, aes(x = surface, y = player0_aces,fill = surface)) +
  geom_boxplot() + scale_fill_manual(values = surface_colors)
  labs(x = "Surface", y = "Player ace",title = "Players Ace's on Different Surfaces" ) +
  theme(legend.position = "none")
```

From the boxplot we can see the player ace per court and the grass courts have the highest number of aces per game and clay court have the least.This indicates that the player has to focus more on aces in the grass court than other surfaces.

```{r}
# Serve percentage difference
tennis_data_clean$first_serve_pct_diff <- (tennis_data_clean$player0_first_serve_won / tennis_data_clean$player0_first_serve_in) - (tennis_data_clean$player1_first_serve_won / tennis_data_clean$player1_first_serve_in)

# tennis_data_clean$second_serve_pct_diff <- (tennis_data_clean$player0_second_serve_won / tennis_data_clean$player0_first_serve_in) - (tennis_data_clean$player1_second_serve_won / tennis_data_clean$player1_first_serve_in)
tennis_data_clean$breakpoint_save_pct_diff <- with(tennis_data_clean, 
    ifelse(player0_breakpoints_faced == 0, 0, player0_breakpoints_saved / player0_breakpoints_faced) -
    ifelse(player1_breakpoints_faced == 0, 0, player1_breakpoints_saved / player1_breakpoints_faced)
)


# Calculate the ratios of service points won to service games held for both players
tennis_data_clean$player0_serve_points_won_to_games_held_ratio <- ifelse(tennis_data_clean$player0_service_games_held == 0, NA, tennis_data_clean$player0_servepointswon / tennis_data_clean$player0_service_games_held)
tennis_data_clean$player1_serve_points_won_to_games_held_ratio <- ifelse(tennis_data_clean$player1_service_games_held == 0, NA, tennis_data_clean$player1_servepointswon / tennis_data_clean$player1_service_games_held)

# Calculate the difference between the two ratios
tennis_data_clean$serve_points_won_to_games_held_diff <- tennis_data_clean$player0_serve_points_won_to_games_held_ratio - tennis_data_clean$player1_serve_points_won_to_games_held_ratio



# Age difference
tennis_data_clean$age_diff <- tennis_data_clean$player0_age - tennis_data_clean$player1_age

# Height difference
tennis_data_clean$height_diff <- tennis_data_clean$player0_height - tennis_data_clean$player1_height

# Rank difference
tennis_data_clean$rank_diff <- tennis_data_clean$player0_rank - tennis_data_clean$player1_rank

# Rank points difference
tennis_data_clean$rank_points_diff <- tennis_data_clean$player0_rank_points - tennis_data_clean$player1_rank_points


```

```{r}
tennis_data_clean <- tennis_data_clean %>%
  select(-player0_age, -player1_age, -player0_height, -player1_height, -player0_first_serve_won, -player0_first_serve_in, -player1_first_serve_won, -player1_first_serve_in, -player0_second_serve_won, -player1_second_serve_won, -player1_rank, -player0_rank, -player0_rank_points, -player1_rank_points,
  -player0_breakpoints_faced,   
  -player0_breakpoints_saved,   
  -player1_breakpoints_faced,   
  -player1_breakpoints_saved,
  -player0_service_games_held,
  -player1_servepointswon,
  -player1_service_games_held,
  -player0_servepointswon,
  -player1_servepointswon,
  -player1_service_games_held,
  -player0_servepointswon,
  -player0_service_games_held,
  -player1_service_games_held,
  -player0_servepointswon,
  -player0_service_games_held,
  -player1_servepointswon,
  -player0_serve_points_won_to_games_held_ratio,
  -player1_serve_points_won_to_games_held_ratio, 
  )
```

```{r}
str(tennis_data_clean)
```

```{r}
numeric_vars <- tennis_data_clean[, sapply(tennis_data_clean, is.numeric)]
numeric_vars
```

```{r}
correlation_matrix <- cor(numeric_vars, use = "complete.obs")
```


```{r}
library(corrplot)
png("heatmap1.png", width = 800, height = 800)
corrplot(correlation_matrix, method = "color", type = "upper", order = "hclust", tl.col = "black", tl.srt = 45)
dev.off()
```


```{r}
threshold <- 0.7
```

```{r}
correlation_matrix_no_diag <- correlation_matrix
diag(correlation_matrix_no_diag) <- 0
```

```{r}
highly_correlated_pairs <- which(abs(correlation_matrix_no_diag) > threshold, arr.ind = TRUE)
highly_correlated_pairs
```

```{r}
highly_correlated_vars <- unique(rownames(highly_correlated_pairs))
highly_correlated_vars
```

```{r}
unique(final_data$best_of)
```

```{r}
str(tennis_data_clean)
```

```{r}
numeric_vars <- sapply(tennis_data_clean, is.numeric)
for (col in colnames(tennis_data_clean)[numeric_vars]) {
  tennis_data_clean[[col]] <- ifelse(is.na(tennis_data_clean[[col]]), mean(tennis_data_clean[[col]], na.rm = TRUE), tennis_data_clean[[col]])
}
```



```{r}
# Drop rows with missing values in player0_hand or player1_hand
tennis_data_clean <- tennis_data_clean[!is.na(tennis_data_clean$player0_hand) & !is.na(tennis_data_clean$player1_hand), ]

```


```{r}
# Analyze missing values in the dataset
missing_values <- tennis_data_clean %>%
  gather(variable, value) %>%
  mutate(is_missing = is.na(value)) %>%
  group_by(variable) %>%
  summarize(missing_count = sum(is_missing), total_count = n(), missing_percentage = missing_count / total_count * 100)

print(missing_values)
```

```{r}
str(tennis_data_clean)
```

```{r}
# Load the required packages
library(caret)
```


```{r}
#tennis_data_clean$winner <- as.factor(tennis_data_clean$winner)
```

```{r}
tennis_data_clean$surface <- as.factor(tennis_data_clean$surface)
tennis_data_clean$tournament_level <- as.factor(tennis_data_clean$tournament_level)
tennis_data_clean$player0_hand <- as.factor(tennis_data_clean$player0_hand)
tennis_data_clean$player1_hand <- as.factor(tennis_data_clean$player1_hand)
```

```{r}
str(tennis_data_clean)
```

```{r}
set.seed(123)
trainIndex <- createDataPartition(tennis_data_clean$winner, p = .7, list = FALSE, times = 1)
train_set <- tennis_data_clean[trainIndex, ]
test_set <- tennis_data_clean[-trainIndex, ]


```

```{r}
logistic_model <- glm(winner ~., family = binomial(link = "logit"), data = train_set)
```

```{r}
predictions <- predict(logistic_model, test_set, type = "response")
predicted_classes <- ifelse(predictions > 0.5, 1, 0)
predicted_classes <- factor(ifelse(predictions > 0.5, 1, 0), levels = c(0, 1))
test_set$winner <- factor(test_set$winner, levels = c(0, 1))
cm <- confusionMatrix(predicted_classes, test_set$winner)
```

```{r}

cm
```

```{r}
cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Prediction", "Reference", "n")

```

```{r}
plot_confusion_matrix <- function(cm_df) {
  ggplot(data = cm_df, aes(x = Reference, y = Prediction, fill = n)) +
    geom_tile() +
    geom_text(aes(label = n), color = "black") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    theme_minimal() +
    xlab("Actual") +
    ylab("Predicted") +
    ggtitle("Confusion Matrix") +
    theme(plot.title = element_text(hjust = 0.5))
}

plot_cm <- plot_confusion_matrix(cm_df)
print(plot_cm)
```
```{r}
accuracy <- cm$overall["Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]
f1_score <- cm$byClass["F1"]

cat("Accuracy: ", accuracy, "\n")
cat("Sensitivity (Recall): ", sensitivity, "\n")
cat("Specificity: ", specificity, "\n")
cat("F1-score: ", f1_score, "\n")

```


```{r}
library(pROC)
roc_curve <- roc(test_set$winner, predictions)
plot(roc_curve, col = "blue", main = "ROC Curve for Logistic Regression Model")
lines(x = c(0, 1), y = c(0, 1), col = "red")
roc_auc <- auc(roc_curve)

```


```{r}
# Load required packages
library(caret)
library(e1071)


# Split data into training and testing sets
ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
svm_model <- train(winner ~ ., data = train_set, method = "svmLinear", trControl = ctrl, tuneGrid = expand.grid(C = c(0.01, 0.1, 1, 10, 100)))
```


```{r}
train_set$winner <- as.factor(train_set$winner)
test_set$winner <- as.factor(test_set$winner)
levels(train_set$winner) <- c("Winner_0", "Winner_1")
levels(test_set$winner) <- c("Winner_0", "Winner_1")
```

```{r}
cv_control <- trainControl(method = "cv", number = 10, savePredictions = "final", classProbs = TRUE, summaryFunction = twoClassSummary)
set.seed(123)
random_forest_model <- train(winner ~ ., data = train_set, method = "rf", trControl = cv_control, metric = "ROC")
print(random_forest_model)
```


```{r}
print(random_forest_model)
```

```{r}
rf_predictions <- predict(random_forest_model, newdata = test_set)
```


```{r}

confusionMatrix(rf_predictions, test_set$winner)

# Calculate accuracy, precision, recall, and F1-score
rf_cm <- confusionMatrix(rf_predictions, test_set$winner)
accuracy <- cm$overall['Accuracy']
precision <- cm$byClass['Pos Pred Value']
recall <- cm$byClass['Sensitivity']
f1_score <- 2 * (precision * recall) / (precision + recall)

cat("Accuracy: ", accuracy, "\n")
cat("Precision: ", precision, "\n")
cat("Recall: ", recall, "\n")
cat("F1-score: ", f1_score, "\n")

```


```{r}

rf_cm_df <- as.data.frame(rf_cm$table)
colnames(rf_cm_df) <- c("Prediction", "Reference", "n")

rf_plot_cm <- plot_confusion_matrix(rf_cm_df)
print(rf_plot_cm)
```


```{r}
# Create the confusion matrix data frame
cm_values <- c(12995, 2352, 2400, 13210)
cm_df <- data.frame(
  Prediction = rep(c("Winner_0", "Winner_1"), each = 2),
  Reference = rep(c("Winner_0", "Winner_1"), times = 2),
  n = cm_values
)

# Use the provided function to plot the confusion matrix
plot_confusion_matrix(cm_df)
```