---
title: "Tennis Match Outcome Predictions"
author: "Team: SNM Analytics"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---



```{r init, include=F}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
options(scientific = T, digits = 3)
# Loading libraries
library(ezids)
library(dplyr)
library(tidyverse)
library(readr)
library(ggplot2)
```

Introduction
In this analysis, we will examine the Jeff Sackmann tennis ATP match dataset to determine the availability of match-level statistics over the years. The dataset contains information on tennis matches from 1968 to 2023, with more extensive match statistics available starting from 1997. Our goal is to visualize the availability of match statistics and explore the data to understand its structure and contents.

```{r}
# Initialize an empty data frame
data <- data.frame()

# Loop through the years and read the corresponding atp_matches files
for (year in 1968:2023) {
  file_path <- paste0("data/atp_matches_", year, ".csv")
  
  if (file.exists(file_path)) {
    temp_data <- read.csv(file_path)
    data <- rbind(data, temp_data)
  }
}
```

```{r}
str(data)
```
```{r}
# Convert tourney_date to a Date object
data$tourney_date <- as.Date(as.character(data$tourney_date), "%Y%m%d")

# Extract year from tourney_date
data$year <- year(data$tourney_date)
```


```{r}
# Select the relevant columns (numeric match statistics)
stats_columns <- c("w_ace", "w_df", "w_svpt", "w_1stIn", "w_1stWon", "w_2ndWon", "w_SvGms", "w_bpSaved", "w_bpFaced",
                   "l_ace", "l_df", "l_svpt", "l_1stIn", "l_1stWon", "l_2ndWon", "l_SvGms", "l_bpSaved", "l_bpFaced")

```

```{r}
# Calculate the percentage of non-null values per year for each numeric variable
non_null_stats <- data %>%
  group_by(year) %>%
  summarize(across(all_of(stats_columns), ~mean(!is.na(.)) * 100, .names = "pct_non_null_{col}"))
```

```{r}
# Gather the data into a long format for easy plotting
non_null_stats_long <- non_null_stats %>%
  gather(key = "variable", value = "pct_non_null", -year)
```

```{r}
# Plot the percentage of non-null values per year for each numeric variable
ggplot(non_null_stats_long, aes(x = year, y = pct_non_null, color = variable)) +
  geom_line() +
  geom_point() +
  labs(title = "Percentage of Non-Null Match Statistics per Year",
       x = "Year",
       y = "Percentage of Non-Null Values",
       color = "Statistic") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 8)) +
  scale_x_continuous(breaks = seq(min(non_null_stats_long$year), max(non_null_stats_long$year), by = 5))


```


```{r}
library(knitr)

kable(non_null_stats, digits = 2)

```

The Jeff Sackman Dataset contains multiple sets of files. We use the Sackman_tennis_atp-master project which contains match level
statistics from the year 1968. Since more extensive match level statistics are only available from the year 1991, we concentrate on the ATP match level files from 1991 to 2023.

```{r}
data <- data %>% filter(year >= 1991)
```


```{r}
str(data)
```

```{r}
# Analyze missing values in the dataset
missing_values <- data %>%
  gather(variable, value, -tourney_id, -tourney_name, -tourney_date, -match_num, -winner_id, -loser_id) %>%
  mutate(is_missing = is.na(value)) %>%
  group_by(variable) %>%
  summarize(missing_count = sum(is_missing), total_count = n(), missing_percentage = missing_count / total_count * 100)

print(missing_values)
```

```{r}
# Analyze missing values by year
missing_values_by_year <- data %>%
  gather(variable, value, -tourney_id, -tourney_name, -tourney_date, -match_num, -winner_id, -loser_id, -year) %>%
  mutate(is_missing = is.na(value)) %>%
  group_by(year, variable) %>%
  summarize(missing_count = sum(is_missing), total_count = n(), missing_percentage = missing_count / total_count * 100) 

print(missing_values_by_year)
```



```{r}
data <- data %>% 
  rename(tournament_id = tourney_id,
         tournament_name = tourney_name,
         tournament_level = tourney_level,
         tournament_date = tourney_date,
         winner_height = winner_ht,
         winner_country = winner_ioc,
         loser_height = loser_ht,
         loser_country = loser_ioc,
         winner_aces = w_ace,
         winner_doublefaults = w_df,
         winner_servepointswon = w_svpt,
         winner_first_serve_in = w_1stIn,
         winner_first_serve_won = w_1stWon,
         winner_second_serve_won = w_2ndWon,
         winner_service_games_held = w_SvGms,
         winner_breakpoints_saved = w_bpSaved,
         winner_breakpoints_faced = w_bpFaced,
         loser_aces = l_ace,
         loser_doublefaults = l_df,
         loser_servepointswon = l_svpt,
         loser_first_serve_in = l_1stIn,
         loser_first_serve_won = l_1stWon,
         loser_second_serve_won = l_2ndWon,
         loser_service_games_held = l_SvGms,
         loser_breakpoints_saved = l_bpSaved,
         loser_breakpoints_faced = l_bpFaced)
```


```{r}
summary(data)
```

```{r}
# Define a custom color palette for the surfaces
surface_colors <- c("Clay" = "#EB9694", 
                    "Carpet" = "#FABE58", 
                    "Grass" = "#87D37C", 
                    "Hard" = "#5DADE2")

# Create the bar chart with custom colors and adjusted font sizes
ggplot(data, aes(x = surface, fill = surface)) + 
  geom_bar(color = "black") +
  scale_fill_manual(values = surface_colors) +
  labs(title = "Total Matches Played on Various Surfaces Since 1991",
       x = "Surface",
       y = "Total Matches") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10), size = 12),
        axis.text = element_text(size = 12),
        panel.grid.major = element_line(colour = "gray90"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 18, hjust = 0.5, color = "gray40"),
        plot.caption = element_text(size = 14, hjust = 0, color = "gray40"),
        plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

```

Most of the matches played are on hard courts!

```{r}
# Filter out rows with missing winner_age and loser_age values
filtered_data <- data %>%
  filter(!is.na(winner_age) & !is.na(loser_age))

# Create a density plot for winner_age and loser_age
ggplot() +
  geom_density(data = filtered_data, aes(x = winner_age, fill = "Winner"), alpha = 0.5) +
  geom_density(data = filtered_data, aes(x = loser_age, fill = "Loser"), alpha = 0.5) +
  labs(x = "Age", y = "Density", title = "Age Distribution of Winners and Losers", fill = "Role") +
  theme_minimal() +
  theme(legend.position = "top")

# Create a box plot for winner_age and loser_age
winner_loser_age <- filtered_data %>%
  select(winner_age, loser_age) %>%
  gather(key = "Role", value = "Age", winner_age, loser_age)

ggplot(winner_loser_age, aes(x = Role, y = Age, fill = Role)) +
  geom_boxplot() +
  labs(x = "", y = "Age", title = "Age Distribution of Winners and Losers") +
  theme_minimal() +
  theme(legend.position = "none")
```
age distribution for "Losers" is wider than "Winners", although mean value is quite the same: younger and olders players are more likely to loose matches.

```{r}

```
