---
title: "Customer Churn Analysis in the Banking Industry"
author: "Team: SNM Analytics"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific = T, digits = 3)
# Loading Required libraries
library(ezids)
library(tidyverse)
library(ggplot2)
```

```{r}
churn_data <- data.frame(read.csv("Data.csv"))
```

```{r}
str(churn_data)
```

```{r}
rows <- nrow(churn_data)
cols <- ncol(churn_data)
```

Our data set has `{r rows}` observations and `{r cols}` variables


```{r}
xkablesummary(churn_data)
```

```{r}
na_count <- colSums(is.na(churn_data))

if (all(na_count == 0)) {
  cat("There are no missing values in the data set.\n")
} else {
  cat("There are some missing values in the data set.\n")
}
```
As We don’t have any missing values in this data set, we don’t need to do anything extra to handle them. 

As defined in our smart question, We only want to look at customers who have a positive balance as they are more valuable customers. We will filter out the data with zero balance.

```{r}
churn_data_filtered <- churn_data %>% filter(Balance > 0)
filtered_rows <- nrow(churn_data_filtered)
```
The filtered data set excluding 0 balance has `{r filtered_rows}` observations



```{r}
# Calculate summary statistics of CreditScore by Exited
churn_credit_score_summary <- churn_data_filtered %>%
  group_by(Exited) %>%
  summarize(
    mean_credit_score = mean(CreditScore),
    sd_credit_score = sd(CreditScore),
    min_credit_score = min(CreditScore),
    max_credit_score = max(CreditScore)
  )

# Print summary table
xkabledply(churn_credit_score_summary)
```

### Churn Data Credit Score Summary

The churn data credit score summary provides a comparison of the credit score statistics between customers who have churned (`Exited = 1`) and customers who have not churned (`Exited = 0`). The summary is calculated on the dataset that includes customers with a balance greater than zero and without removing any outliers. The summary consists of the following columns:

- `Exited`: Churn status (0 for not churned, 1 for churned)
- `mean_credit_score`: Mean credit score for each group
- `sd_credit_score`: Standard deviation of credit scores for each group
- `min_credit_score`: Minimum credit score for each group
- `max_credit_score`: Maximum credit score for each group

The summary table shows that the mean credit score is slightly higher for customers who have not churned (653) compared to customers who have churned (646). The standard deviation for credit scores is also slightly lower for customers who have not churned (95.8) than for customers who have churned (100). The minimum and maximum credit scores for both groups are quite similar, with the lowest credit score for the churned group being 350 and the highest credit score for both groups being 850. This suggests that there is some difference in the variability of credit scores between the two groups of customers. However, this difference may not be very significant or meaningful, as both groups have relatively high standard deviations compared to their mean credit scores. We will perform further analysis or use other measures to better understand how credit score affects customer churn.

Lets also visualize the same on graphs. 


```{r}
ggplot(churn_data_filtered, aes(x = factor(Exited), y = CreditScore, fill = factor(Exited))) +
  geom_boxplot() +
  labs(x = "Churn Status (0 = Not Churned, 1 = Churned)", y = "Credit Score", fill = "Churn Status") +
  theme_minimal() +
  ggtitle("Credit Score Distribution by Churn Status") +
  scale_fill_manual(values = c("0" = "lightgreen", "1" = "indianred1"), labels = c("0" = "Not Churned (0)", "1" = "Churned (1)"))
```

### Box Plot of Credit Score Distribution by Churn Status

The box plot for the given data displays the distribution of credit scores by churn status. It has two box plots, one for customers who have not churned (`Exited = 0`) and one for customers who have churned (`Exited = 1`).

Each box plot has the following key components:

- **Median** (the horizontal line inside the box): This represents the middle value of the credit scores in each group.
- **Lower quartile** (the left edge of the box): This represents the value below which 25% of the credit scores in each group lie.
- **Upper quartile** (the right edge of the box): This represents the value below which 75% of the credit scores in each group lie.
- **Interquartile range** (IQR, the width of the box): This represents the range within which the middle 50% of the credit scores in each group lie.
- **Whiskers** (the lines extending from the box): These represent the range of credit scores that are within 1.5 times the IQR from the lower and upper quartiles.
- **Outliers** (individual points outside the whiskers): These represent credit scores that fall outside the whisker range and are considered unusual or extreme values.

Based on the box plot, we can make the following observations:

- The median credit score for both groups (churned and not churned) is quite similar(around 650), indicating that there might not be a significant difference in *the central tendency* of credit scores between
the two groups.

- The overall range of credit scores and *the interquartile ranges* for both groups are also quite similar, suggesting that *the spread* of credit scores is not considerably different between the churned and not churned groups.

- There might be a few outliers in both groups, but their presence seems to be more visible in churned customers.



```{r}
ggplot(churn_data_filtered, aes(x = CreditScore, fill = factor(Exited))) +
  geom_density(alpha = 0.5) +
  labs(x = "Credit Score", y = "Density", fill = "Churn Status") +
  theme_minimal() +
  ggtitle("Credit Score Distribution Density by Churn Status") +
  scale_fill_manual(values = c("0" = "lightgreen", "1" = "indianred1"), labels = c("0" = "(0 = Not Churned)", "1" = "(1 = Churned)"))
```


```{r}
ggplot(churn_data_filtered, aes(x = CreditScore)) +
  geom_histogram(colour="black",fill = "lightblue") +
  labs(title = "Credit Score Countplot", x = "Credit Score", y = "Count")
```
This is a Credit Score distribution for customers.

```{r}
ggplot(churn_data_filtered, aes(x = Geography)) +
  geom_bar(fill = c("lightblue", "indianred", "lightgreen")) +
  labs(title = "Customer vs Country", x = "Geography", y = "Count")
```

Most of the customers with a account balance greater than 0 orginate from france and germany which is double that of spain.

```{r}
ggplot(churn_data_filtered, aes(x = NumOfProducts)) +
  geom_bar(fill = "lightgrey") +
  labs(title = "Products owned by Customers", x = "products", y = "Count")
```

We can see that the majority of customers have 1 or 2 products, while relatively few customers have 3 or 4 products with a balance greater than 0.

